Convergent Evolution During Local Adaptation to Patchy Landscapes

### 2015-10-19 patchy-selection-paper.tex

%%%%%%%%%%%
\section*{Abstract}
Species often encounter, and adapt to, many patches of locally similar environmental conditions across their range. 
Such adaptation can occur through convergent evolution if different alleles arise and spread in different patches, 
or through the spread of shared alleles by migration
acting to synchronize adaptation across the species.
The tension between the two reflects 
the degree of constraint imposed on evolution by the underlying genetic architecture 
versus how effectively selection and geographic isolation act to inhibit the geographic spread of locally adapted alleles.
This paper studies a model of the balance between these two routes to adaptation in continuous environments with patchy selection pressures.
We address the following questions: 
How long does it take for a novel, locally adapted allele to appear in a
patch of habitat where it is favored through mutation?
Or, through migration from another, already adapted patch?
Which is more likely to occur, as a function of distance between the patches?
How can we tell which has occurred, i.e., what population genetic
signal is left by the spread of migrant alleles?
To answer these questions we examine the family structure underlying migration--selection equilibrium
surrounding an already adapted patch,
in particular treating those rare families that reach new patches
as spatial branching processes.
A main result is that locally adapted alleles in different patches
are likely to be of independent origin
if the distance between the patches is larger than
$(\sigma/s_m)\log(Cs_m/\mu)$, 
where $\sigma$ is the dispersal distance, 
$s_m$ is the selective disadvantage of these alleles between patches,
$\mu$ is the mutation rate, and $C$ is a constant depending on the probability of establishment and patch geometry.
This provides a way to understand the role of geographic separation between patches 
in promoting convergent adaptation and the genomic signals it leaves behind.
We illustrate these ideas using the convergent evolution of cryptic coloration in the rock pocket mouse, \textit{Chaetodipus intermedius}, as an empirical example.

>>>>

%%%%%%%%%%%
\section*{Abstract}
Species often encounter, and adapt to, many patches of locally similar environmental conditions across their range. 
Such adaptation can occur through convergent evolution if different alleles arise and spread in different patches, 
or through the spread of shared alleles by migration
acting to synchronize adaptation across the species.
The tension between the two reflects 
the degree of constraint imposed on evolution by the underlying genetic architecture 
versus how effectively selection and geographic isolation act to inhibit the geographic spread of locally adapted alleles.
This paper studies a model of the balance between these two routes to adaptation in continuous environments with patchy selection pressures.
We address the following questions: 
How long does it take for a novel, locally adapted allele to appear in a
patch of habitat where it is favored through mutation?
Or, through migration from another, already adapted patch?
Which is more likely to occur, as a function of distance between the patches?
How can we tell which has occurred, i.e., what population genetic
signal is left by the spread of migrant alleles?
To answer these questions we examine the family structure underlying migration--selection equilibrium
surrounding an already adapted patch,
in particular treating those rare families that reach new patches
as spatial branching processes.
A main result is that patches further apart than a critical distance will
likely evolve independent locally adapted alleles;
this distance is proportional to the spatial scale of selection
($\sigma/\sqrt{s_m}$, where $\sigma$ is the dispersal distance and
$s_m$ is the selective disadvantage of these alleles between patches),
and depends linearly on $\log(s_m/\mu)$, where $\mu$ is the mutation rate.
This provides a way to understand the role of geographic separation between patches 
in promoting convergent adaptation and the genomic signals it leaves behind.
We illustrate these ideas using the convergent evolution of cryptic coloration in the rock pocket mouse, \textit{Chaetodipus intermedius}, as an empirical example.
